\documentclass[10pt,twoside,a4paper]{report}

\usepackage{a4}

\setlength{\textwidth}{5.5in}
\setlength{\oddsidemargin}{0.77in}
\setlength{\evensidemargin}{0in}
\setlength{\topmargin}{0in}

\usepackage{url}

\usepackage[
    plainpages=false,
    pdfpagelabels,
    bookmarksnumbered,
    hyperindex=true
]{hyperref}

\usepackage[usenames,dvipsnames]{color}

\hypersetup{
    bookmarks=true,         % show bookmarks bar?
    unicode=false,          % non-Latin characters in Acrobat’s bookmarks
    pdftoolbar=true,        % show Acrobat’s toolbar?
    pdfmenubar=true,        % show Acrobat’s menu?
    pdffitwindow=false,     % window fit to page when opened
    pdfstartview={FitH},    % fits the width of the page to the window
    pdftitle={My title},    % title
    pdfauthor={Author},     % author
    pdfsubject={Subject},   % subject of the document
    pdfcreator={Creator},   % creator of the document
    pdfproducer={Producer}, % producer of the document
    pdfkeywords={keywords}, % list of keywords
    pdfnewwindow=true,      % links in new window
    colorlinks=true,        % false: boxed links; true: colored links
    linkcolor=red,          % color of internal links
    citecolor=green,        % color of links to bibliography
    filecolor=magenta,      % color of file links
    urlcolor=cyan           % color of external links
}

\usepackage{listings}

% \lstset{% general command to set parameter(s)
%     language=C++,
%     basicstyle=\small\ttfamily, % print whole listing small
%     % basicstyle=\footnotesize\ttfamily,
%     keywordstyle=\bfseries\color{OliveGreen},
%     commentstyle=\itshape\color{Purple},
%     identifierstyle=\color{blue},
%     stringstyle=\color{Orange},
%     showstringspaces=false,      % no special string spaces
%     captionpos=b
% }
\lstset{% general command to set parameter(s)
    language=C++,
    basicstyle=\ttfamily\small,
    identifierstyle=\sffamily\color{blue},
    keywordstyle=\sffamily\bfseries\color{OliveGreen},,
    commentstyle=\rmfamily\color{Purple},
    stringstyle=\rmfamily\itshape\color{Orange},
    numberstyle=\scriptsize,
    showstringspaces=false,
    showstringspaces=false,      % no special string spaces
    captionpos=b
}

\newcommand{\includecode}[3] { %
\lstinputlisting[
label=#2,
firstline=\csname #2firstline\endcsname,
lastline=\csname #2lastline\endcsname,
float,
caption=#3]
{../Src/#1}
}

\begin{document}

\title{The Kamin Interpreters in C++}
\author{Tim Budd}
\maketitle
\cleardoublepage

\begin{abstract}
    This paper describes a series of interpreters for the languages used in the
    book ``Programming Languages: An Interpreter-Based Approach'' by Samuel
    Kamin (Addison-Wesley, 1989).  Unlike the interpreters provided by Kamin,
    which are written in Pascal, these interpreters are written in C++.  It is
    my belief that the use of inheritance in C++ better illustrates the unique
    features of each of the several languages.  In the Pascal versions of the
    interpreters the differences between the various interpreters, although
    small, are scattered throughout the code.  In the C++ versions differences
    are produced using only the mechanism of subclassing.  This means that the
    vast majority of code remains the same, and differences can be much more
    precisely isolated.

    The chapters in this report correspond to the chapters in the original text.
    Where motivational or background material is provided in that source it is
    generally omitted here.  A major exception is in those places (chiefly
    chapters 3, 7 and 8) where I have selected a syntax slightly different from
    that provided by Kamin.

    The use of an Object-Oriented language for the interpreters may seem a bit
    incongruous, since Object-Oriented programming is not discussed until
    Chapter 7.  Nevertheless, I think the benefits of programming the
    interpreters in C++ outweighs this problem.
\end{abstract}

\tableofcontents
\listoffigures
\lstlistoflistings

\include{listingLines}
\include{chap1}
\include{chap2}
\include{chap3}
\include{chap4}
\include{chap5}
\include{chap6}
\include{chap7}
\include{chap8}
\include{future}

\end{document}
