\chapter{The Smalltalk interpreter}

As with chapter 3, with the Smalltalk interpreter I have also made a number of
changes.  These include the following:

\begin{itemize}
    \item I have changed the syntax for message passing.  The first argument in
    a message passing expression is an object, which is defined (for
    implementation purposes) as a type of function.  The second argument must be
    the message selector, a symbol.  This change is not only produces a syntax
    that is slightly more Smalltalk-like, but it more closely reinforces the
    critical object-oriented idea that the interpretation of a message depends
    upon the receiver for that message.
    \item Integers are objects, and respond to messages.  The most obvious
    effect of this is to restore infix syntax for arithmetic operations, since
    (3 + 4) is interpreted (Smalltalk-like) as the message ``+'' being passed to
    the object 3 with argument 4.
    \item The initial environment is very spare.  There are only the two classes
    {\sf Object} and {\sf Integer}, which respond to the messages {\sf
        subclass}, {\sf method} and {\sf new}, and integer instances that
    respond to arithmetic messages.
    \item The {\sf if} command is a message sent to integers (0 for false and
    non-zero for true).  This is also more Smalltalk-like.  The following
    expression sets {\sf z} to the minimum of {\sf x} and {\sf y}.
    \begin{center} {\sf ((x $<$ y) if (set z x) (set z y))}
    \end{center}
    \item The only non-message statements are the assignment statement {\sf set}
    and the {\sf begin} statement.  (Note - there is no loop.  I couldn't think
    of a good way to do this within the syntax given using message passing (no
    blocks!) but I don't think this will be too great a problem; recursion can
    be used in most cases where looping is used currently).
\end{itemize}

A class hierarchy for the classes added in this chapter is shown in
Figure~\ref{smallhier}.

\setlength{\unitlength}{5mm}
\begin{figure}
    \begin{picture}(25,10)(-4,-5)
        \put(-3.5,0){\sf Expression}
        \put(0,0.2){\line(1,0){1}}
        \put(1,0){\sf Function}
        \put(0,0.2){\line(1,4){1}}
        \put(1,4){\sf Symbol}
        \put(3,4.2){\line(1,0){1}}
        \put(4,4){\sf SmalltalkSymbol}
        \put(4,0.2){\line(1,-2){1}}
        \put(5,-2){\sf UserFunction}
        \put(9,-1.8){\line(1,0){1}}
        \put(10,-2){\sf method}
        \put(13,-1.8){\line(1,0){1}}
        \put(14,-2){\sf SubclassMethod}
        \put(13,-1.8){\line(1,1){1}}
        \put(14,-1){\sf NewMethod}
        \put(13,-1.8){\line(1,2){1}}
        \put(14,0){\sf IntegerBinaryMethod}
        \put(13,-1.8){\line(1,-1){1}}
        \put(14,-3){\sf IfMethod}
        \put(13,-1.8){\line(1,-2){1}}
        \put(14,-4){\sf MethodMethod}
        \put(4,0.2){\line(1,2){1}}
        \put(5,2){\sf Object}
        \put(7,2.2){\line(1,0){1}}
        \put(8,2){\sf IntegerObject}
    \end{picture}
    \caption{Expression class hierarchy for Smalltalk interpreter}
    \label{smallhier}
\end{figure}

\section{Objects and Methods}

An object is an encapsulation of behavior and state.  That is, an object
maintains, like a cluster, certain state information accessible only within the
object.  Similarly objects maintain a collection of functions, called {\em
    methods}, that can be invoked only via message passing.  Internally, both
these are represented by environments (Figure~\ref{SmalltalkObject}).  The
methods environment contains a collection of functions, and the data environment
contains a collection of internal variables.  Objects are declared as a subclass
of {\sf function} so that normal function syntax can be used for message
passing.  That is, a message is written as

\begin{center}
{\sf (object message arguments)}
\end{center}
%
\includecode{smalltalk.C}{SmalltalkObject}
{Classes for Object and Method}

Methods are similar to conventional functions (and are thus subclasses of {\sf
    UserFunction}) in that they have an argument list and body.  Unlike
conventional functions they have a receiver (which must always be an object) and
the environment in which the method was created, as well as the environment in
which the method is invoked.  Thus methods define a new message {\sf doMethod}
that takes these additional arguments.

A subtle point to note is that the creation environment in normal functions is
captured when the function is defined.  For objects this environment cannot be
defined when the methods are created, but must wait until a new instance is
created.  Our implementation waits even longer, and passes it as part of the
message passing protocol.

The mechanism of message passing is defined by the function {\sf apply} in class
{\sf Object} (Figure~\ref{SmalltalkObjectApply}).  Messages require a symbol for
the first argument, which must match a method for the object.  This method is
then invoked.  Similarly Figure~\ref{SmalltalkObjectApply} shows the execution
of normal methods (that is, those methods other than the ones provided by the
system).  The execution context is set for the method, and the receiver is added
as an implicit first argument, called {\sf self} in every method.  The method is
then invoked as if it were a conventional function.
%
\includecode{smalltalk.C}{SmalltalkObjectApply}
{Implementation of Message Passing}

\section{Classes}

Classes are simply objects.  As such, they respond to certain messages.  In our
Smalltalk interpreter there are initially two classes, {\sf Object} and {\sf
    Integer}.  The class {\sf Object} is a superclass of {\sf Integer}, and is
typically the superclass of most user defined classes as well.  There are
initially three messages that classes respond to:

\begin{itemize}
    \item {\sf subclass}.  This message is used to create new classes, as
    subclasses of existing classes.  Any arguments provided are treated as the
    names of instance variables (local state) to be generated when instances of
    the new classes are created.  The new class is returned as an object, and is
    usually immediately assigned to a global variable.  The syntax for new
    classes is thus similar to the following:
    \begin{center}
        {\sf (set Foo (Object subclass x y z))}
    \end{center}
    which creates a new class with three instance variables, and assigns this
    class to the variable {\sf Foo}.  Subclasses can also access instance
    variables defined in classes.

    It is legal to subclass from class {\sf Integer}, although the results are
    not useful for any purpose.

    \item {\sf new}.  This message, which takes no arguments, is used to create
    a new instance of the receiver class.  The new instance is returned as the
    result of the method, as in the following:
    \begin{center}
        {\sf (set newfoo (Foo new))}
    \end{center}
    Although the class {\sf Integer} responds to the message {\sf new}, no
    useful value is returned.  (Real Smalltalk has something called {\em
        metaclasses} that can be used to prevent certain classes from responding
    to all messages.  Our Smalltalk doesn't).
    \item {\sf method}.  This message is used to define a new method for a
    class.  Following the keyword {\sf method} the syantx is the same as a
    normal function definition.  Within a method the pseudo-variable {\sf self}
    can be used to represent the receiver for the method.
    \begin{center}
        {\sf (Integer method square () (self * self))}
    \end{center}
\end{itemize}

Classes are represented in the same format as other objects.  They act as if
they held two instance variables; {\sf names}, which contains a list of instance
variable names for the class, and {\sf methods}, which contains the table of
method definitions for the class.  Note that these are held in the data area for
the class.  (A picture might help here...).

The implementation of the method {\sf subclass} is shown in
Figure~\ref{SmalltalkSubclassMethod}.  The instance variables for the parent
class is obtained, and the new instance variables for the class added to them.
Inheritance is implemented by creating a new empty method table, but having it
point to the method table for the parent class.  Thus a search of the method
table for the newly created class will automatically search the parent class if
no overriding method is found.  These two values are inserted as data values in
the new class object.  The methods a class responds to will be exactly the same
as those of the parent class (thus all classes respond to the same messages).
%
\includecode{smalltalk.C}{SmalltalkSubclassMethod}
{Implementation of the {\sf subclass} method}
%
\includecode{smalltalk.C}{SmalltalkObjectGetNames}
{Implementation of Object getNames}

The implementation of the method {\sf new}, shown in
Figure~\ref{SmalltalkNewMethodDoMethod}, gets the list of instance variables
associated with the class.  A new environment is then created that assigns an
empty value to each variable.  Using the method table stored in the data area
for the class object a new object is then created.
%
\includecode{smalltalk.C}{SmalltalkNewMethodDoMethod}
{The method {\sf new}}
%
\includecode{smalltalk.C}{SmalltalkMethodMethodDoMethod}
{The method {\sf method}}

The method used to respond to the {\sf method} command is shown in
Figure~\ref{SmalltalkMethodMethodDoMethod}.  This is very similar to the
function used to break apart the {\sf define} command in Chapter 1.  The only
significant difference includes the addition of the receiver {\sf self} as an
implicit first parameter in the argument list, and the fact that the function is
placed in a method table, rather than in the global environment.

\section{Symbols and Integers}

Symbols in Smalltalk have no property other than they evaluate to themselves,
and are guaranteed unique.  They are easily implemented by subclassing the
existing class {\sf Symbol} (Figure~\ref{SmalltalkSymbol}), and modifying the
reader/parser to recognize the tokens.  (Unlike symbols in real Smalltalk, our
symbols are not objects and will not respond to any messages).
%
\includecode{smalltalk.C}{SmalltalkSymbol}
{Symbols in Smalltalk}
%
\includecode{smalltalk.C}{SmalltalkInteger}
{Integers in Smalltalk}

Integers are also redefined as objects, and a built-in method {\sf
    IntegerBinaryMethod} (Figure~\ref{SmalltalkInteger}), similarly to {\sf
    IntegerBinaryFunction}, is created to simplify the arithmetic methods.

Control flow is implemented as a message to integers.  (In real Smalltalk
control flow is implemented as messages, but to different objects).  If the
receiver is zero the first argument to the if method is returned, otherwise the
second argument is returned.
%
\includecode{smalltalk.C}{SmalltalkIfMethod}
{Implementation of the if method}

\section{Smalltalk reader}

The Smalltalk reader subclasses the reader class so as to recognize
integers and symbols (Figure~\ref{SmalltalkReader}).
%
\includecode{smalltalk.C}{SmalltalkReader}
{The Smalltalk reader}

\section{The big bang}

To initialize the interpreter we must create the objects {\sf Object} and {\sf
    Integer}.  (Need more explanation here, but I'll just give the code for
now).
%
\includecode{smalltalk.C}{SmalltalkInitialize}
{Initializing the Smalltalk interpreter}
